---
title: "High value customers identification using K-Means clustering"
author: "Sayorn Chin"
date: "06/10/2021"
output: rmarkdown::github_document
---

### 1. Problem statement
A UK-based online retail store has captured the sales data for different products for the period of one year (Nov 2016 to Dec 2017). The organization sells gifts primarily on the online platform. The customers who make a purchase consume directly for themselves. There are small businesses that buy in bulk and sell to other customers through the retail outlet channel.

### 2. Project objective
This project aims to find significant customers for the business who make high purchases of their favourite products. The organization wants to roll out a loyalty program to the high-value customers after identification of segments. I use the clustering methodology to segment customers into groups.

### 3. Data and method
#### 3.1 Data
```{r, warning=FALSE, message=FALSE}
## Required R libraries
load_libraries <- function(){
  library(usdm)
  library(sp)
  library(raster)
  library(dplyr)
  library(ggplot2)
  library(cluster)
  library(purrr)
  library(tidyverse)
  library(factoextra)
  library(DataExplorer)
  library(gridExtra)
  library(lubridate)
  library(NbClust)
  library(countrycode)
print("The libraries are loaded.")
}
load_libraries()

## Set working directory
setwd("/Users/schinlfc/data-science-R/e-commerce-clustering/data")

## Load dataset
ecommerce <- read.csv("Ecommerce.csv")

## Data preprocessing
# View the structure of the data frame
str(ecommerce)
# View basic summary statistics
summary(ecommerce)
# View number of rows and columns
dim(ecommerce)
# Find the number of missing values in the rows and columns
namat <- is.na(ecommerce)
table(rowSums(namat))
# Find the total number of missing values in each column
colSums(namat)
# The alternative way to find the the number of missing values in each column
naecommerce <- ecommerce %>% mutate_all(is.na)
print(naecommerce %>% summarise_all(sum))
# Remove column 'X' since the entire column fill with NA
ecommerce <- ecommerce[, -9]
# Remove missing rows for column 'CustomerID' 
ecommerce <- na.omit(ecommerce)
# Check if any missing NAs remain
sum(!complete.cases(ecommerce))
# View number of rows after removing NA's
nrow(ecommerce)
# Convert column 'InvoiceDate' to appropriate date class
ecommerce <- ecommerce %>% mutate(InvoiceDate = lubridate::dmy(InvoiceDate))
head(ecommerce, 10)
# Create a 'Revenue' column
ecommerce <- ecommerce %>% mutate(Revenue = UnitPrice * Quantity)
head(ecommerce, 10)

## Data exploration
# Plotting total revenue by invoice date
ecommerce %>% 
  group_by(InvoiceDate) %>% 
  summarise(total_revenue = sum(Revenue)) %>%
  ggplot(aes(InvoiceDate, total_revenue)) +
    geom_point(color="blue") +
    geom_smooth(method="auto", se=TRUE, color="red", fill="red",
                fullrange=FALSE, level=0.95) +
    labs(title = "Total revenue by invoice date") +
    xlab("Invoice date") +
    ylab("Total revenue")
# Generate a table of total revenue and total quantities sold by 'CustomerID'
ecommerce %>% group_by(CustomerID) %>% 
  summarise(total_revenue = sum(Revenue), 
            total_qunt_sold = sum(Quantity)) 
# Why CustomerID == '12346' shows zero for total revenue and total quantities sold?
ecommerce %>% filter(CustomerID == '12346') %>% 
  select (CustomerID, Revenue, Quantity)
# Generate a table of total revenue and otal quantities sold by 'Country'
ecommerce %>% group_by(Country) %>% 
  summarise(total_revenue = sum(Revenue), 
            total_quant_sold = sum(Quantity))
# Generate a table of total revenue and otal quantities sold by 'Country'
ecommerce %>% group_by(InvoiceDate, CustomerID, Country) %>% 
  summarise(total_revenue = sum(Revenue), 
            total_quant_sold = sum(Quantity))
# Generate a table of total revenue and otal quantities sold by 'InvoiceDate, CustomerID, and Country'
cust_data <- ecommerce %>% 
  group_by(CustomerID, Country) %>% 
  summarise(total_revenue = sum(Revenue), 
            total_quant_sold = sum(Quantity))
# Show the first 10 rows
head(cust_data, 10)
# Show the number of customers
nrow(cust_data)
# Show the exact number of customers
length(unique(cust_data$CustomerID))
# Remove customers who lived in multiple countries in the same year
freq_table <- data.frame(table(cust_data$CustomerID))
head(freq_table, 10)
uniqueCustomerID = (freq_table[freq_table$Freq==1,])$Var1
head(uniqueCustomerID, 10)
cust_data <- subset(cust_data, (cust_data$CustomerID) %in%
                    uniqueCustomerID)
head(cust_data, 10)
# Show whether duplicated customers are successfully removed
nrow(cust_data)
length(unique(cust_data$CustomerID))
# Plot to see whether there are outliers in 'total_revenue' and 'total_quant_sold' columns
ggplot(cust_data, aes(total_revenue)) +
  geom_boxplot(fill = 'white', colour= '#3366FF', notch = TRUE,
               outlier.colour = "red", outlier.shape = 2) +
  xlab('Total revenue')

ggplot(cust_data, aes(total_quant_sold)) +
  geom_boxplot(fill = 'white', colour= '#3366FF', notch = TRUE,
               outlier.colour = "red", outlier.shape = 2) +
  xlab('Total quantities sold')

## Remove outliers
# Clearly there are outliers for 'total_revenue' and 'total_quant_sold' columns, so I would adopt the IQR method to remove these outliers because it does not depend on the mean and standard deviation of the dataset. Note: the interquartile range is the central 50% or the area between the 75th and the 25th percentile of a distrubtion. A poin is an outlier iif it is above the 75th or below the 25th percentile by a factor of 1.5 times the IQR. For example, if Q1 = 25th percentile and Q3 = 75th percentile, then IQR = Q3 - Q1. And an outlier would be a point below [Q1 - (1.5) * IQR] or above [Q3 + (1.5)I * QR]
Q <- quantile(cust_data$total_revenue, 
              probs = c(.25, .75), na.rm = FALSE)
iqr <- IQR(cust_data$total_revenue)
# Find the cut-off ranges beyond which all data points are outliers
up <- Q[2] + 1.5 * iqr # Upper Range
low <- Q[1] - 1.5 * iqr # Lower Range
# Eliminating outliers
eliminated <- subset(cust_data, cust_data$total_revenue > low & cust_data$total_revenue < up)
# Plot 'total_revenue' column without outliers
ggplot(eliminated, aes(total_revenue)) +
  geom_boxplot(fill = 'white', colour= '#3366FF', notch = TRUE,
               outlier.colour = "red", outlier.shape = 2) +
  xlab('Total revenue')
# Remove outliers from 'total_quant_sold' column
Q1 <- quantile(eliminated$total_quant_sold,
              probs = c(.25, .75), na.rm = FALSE)
iqr1 <- IQR(eliminated$total_quant_sold)
# Find the cut-off ranges beyond which all data points are outliers
up1 <- Q1[2] + 1.5 * iqr # Upper Range
low1 <- Q1[1] - 1.5 * iqr # Lower Range
# Eliminating outliers
cust_data <- subset(eliminated, eliminated$total_quant_sold > low & eliminated$total_quant_sold < up)
# Plot 'total_revenue' column without outliers
ggplot(cust_data, aes(total_quant_sold)) +
  geom_boxplot(fill = 'white', colour= '#3366FF', notch = TRUE,
               outlier.colour = "red", outlier.shape = 2) +
  xlab('Total quantities sold')

## Remove 'CustomerID' and 'Country' columns
cust_data_new <- cust_data[, -c(1, 2)]
head(cust_data_new, n=10)

## Scaling
# Before proceed to model building, scaling is an important step.
cust_data_new <- scale(cust_data_new)
head(cust_data_new, n=10)
```

#### 3.2 Method
##### 3.2.1 K-Means clustering
```{r, warning=FALSE, message=FALSE}
set.seed(874)
# Elbow method
fviz_nbclust(cust_data_new, kmeans, method = "wss") +
  geom_vline(xintercept = 2, linetype = 2)+
  labs(subtitle = "Elbow method")

# Silhouette method
set.seed(2345) 
fviz_nbclust(cust_data_new, kmeans, method = "silhouette")+
  labs(subtitle = "Silhouette method")
 
# Gap statistic
# nboot = 100 to keep the function speedy. 
# Recommended value: nboot= 500 for your analysis.
# Use verbose = FALSE to hide computing progression.
set.seed(123)
fviz_nbclust(cust_data_new, kmeans, nstart = 25,  method = "gap_stat", nboot = 100, verbose = FALSE)+
  labs(subtitle = "Gap statistic method")

# Perform k-means clustering based on the methods above with k=2
km <- kmeans(cust_data_new, centers = 2, iter.max = 50, nstart = 50)
str(km)
km$centers
# Visualize the k-means
fviz_cluster(km, cust_data_new,
             palette = c("#2E9FDF", "#00AFBB"), 
             geom = "point",
             ellipse.type = "convex",
             ylab = "Total quantities sold",
             xlab = "Total revenue"
             )
```

### 4. Results
```{r, warning=FALSE, message=FALSE}
# Assign cluster to 'cust_data' data frame to identify valuable customers
cust_data$cluster <- km$cluster
head(cust_data, n=10)

# Plot count of clusters by country
cust_data %>% mutate(Country = ifelse(Country == 'RSA', 'South Africa', Country)) %>%
ggplot( aes(y = cluster)) +
  geom_bar(aes(fill = Country)) +
  ggtitle("Count of clusters by country") +
  ylab("Cluster") +
  xlab("Count") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot count of clusters by country without United Kingdom
cust_data %>% filter(Country!= 'United Kingdom') %>%
ggplot( aes(y = cluster)) +
  geom_bar(aes(fill = Country)) +
  ggtitle("Count of clusters by country without United Kingdom") +
  ylab("Cluster") +
  xlab("Count") +
  theme(plot.title = element_text(hjust = 0.5))

# Create 'region' column using library(countrycode)
cust_data <- cust_data %>% 
  mutate(Region = countrycode(sourcevar = Country,
                                          origin = "country.name",
                                          destination = "region"))
# Remove unidentified region
cust_data <- na.omit(cust_data)
  
# Plot count of clusters by region
ggplot(data = cust_data, aes(y = cluster)) +
  geom_bar(aes(fill = Region)) +
  ggtitle("Count of clusters by region") +
  ylab("Cluster") +
  xlab("Count") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot count of clusters by region excluding United Kingdom
cust_data %>% filter(Country!= 'United Kingdom') %>%
  ggplot( aes(y = cluster)) +
    geom_bar(aes(fill = Region)) +
    ggtitle("Count of clusters by region without United Kingdom") +
    ylab("Cluster") +
    xlab("Count") +
    theme(plot.title = element_text(hjust = 0.5))
```

### 5. Conclusion
This project aims to identify valuable customers for a UK-based online retail store. By adopting K-Means clustering, I managed to segment customers into two groups. I defined valuable customers by constructing total revenue generated by the store over time, and total quantities sold. With these variables and the K-Means clustering technique, I was able to breaking down the count of each cluster by country and region. For future, work it would be ideal to break down the count of each cluster by age, gender, and education if data were available. This will give this the organization a better measure in terms of the roll out of the loyalty program to the high-value customers.   

